The Calculus of Linear Constructions

### 2021-10-10 writing/draft1.tex
draft1

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{10pt}
\setlength{\grammarindent}{10em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\LNLD}{\text{LNL}_\text{D}}
\newcommand{\ok}{\textbf{ok}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{$\text{let }#1\text{ := }#2\text{ in }#3$}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\subst}{\text{subst}}
\newcommand{\utype}{:_U}
\newcommand{\ltype}{:_L}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}

@@

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encodes the usage of resource. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$, and a linear arrow $\multimap_L$ as follows.
  \begin{equation*}
    Money\ 5 \multimap_L Camels + Marlboro
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x : A) \rightarrow_s B$ | $A \rightarrow_s B$
      \indalt $(x : A) \multimap_s B$ | $A \multimap_s B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory is the presence of four function types: $(x : A) \rightarrow_s B$, $A \rightarrow_s B$, $(x : A) \multimap_s B$, $A \multimap_s B$. The reason for these variants is that we have built the ! exponential of linear logic directly into the syntax of function types. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into function syntax, we implicitly limit ! to only be used canonically. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing extra machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ : A) \rightarrow_U B \quad &\equiv \quad !(!A \multimap !B) \\
      (\_ : A) \rightarrow_L B \quad &\equiv \quad !(!A \multimap B) \\
      A \rightarrow_U B \quad &\equiv \quad !(A \multimap !B) \\
      A \rightarrow_L B \quad &\equiv \quad !(A \multimap B) \\
      (\_ : A) \multimap_U B \quad &\equiv \quad !A \multimap !B \\
      (\_ : A) \multimap_L B \quad &\equiv \quad !A \multimap B \\
      A \multimap_U B \quad &\equiv \quad A \multimap !B \\
      A \multimap_L B \quad &\equiv \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (2), (5), (6) if the co-domain $B$ is dependent on arguments of the domain $A$. We will discuss the detail of these function types in Section \ref{tyformation}.

@@

  \begin{definition}
    The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
    \begin{enumerate}
      \item $\preceq$ is a partial order with respect to definitional equality.
        \begin{enumerate}
          \item If $A \equiv B$, then $A \preceq B$.
          \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
          \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
        \end{enumerate}
      \item $U_* \preceq U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
      \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        \begin{enumerate}
          \item $(x : A_1) \rightarrow B_1 \preceq (x : A_2) \rightarrow B_2$
          \item $(x : A_1) \multimap B_1 \preceq (x : A_2) \multimap B_2$
          \item $A_1 \rightarrow B_1 \preceq A_2 \rightarrow B_2$
          \item $A_1 \multimap B_1 \preceq A_2 \multimap B_2$
        \end{enumerate}
    \end{enumerate}
  \end{definition}

  Figure \ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_k$ has $U_{k+1}$ as its type, allowing functions to dependently quantify over linear \textit{types}. However, $L_k$ cumulates to $L_{k+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universe, and subsequently losing track of its occupants' linearity.

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype U_i \\ 
        \Gamma, x \utype A \vdash B \utype U_* }
      { \Gamma \vdash (x : A) \rightarrow_U B \utype U_* } 
      \rname{U-Prop}
      \\

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype U_k \\ 
        \Gamma, x \utype A \vdash B \utype s_k }
      { \Gamma \vdash (x : A) \rightarrow_s B \utype U_k } 
      \rname{U-Prod}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype L_k \\ 
        \Gamma \vdash B \utype s_k }
      { \Gamma \vdash A \rightarrow_s B \utype U_k } 
      \rname{Arrow}
      \\

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype U_k \\ 
        \Gamma, x \utype A \vdash B \utype s_k }
      { \Gamma \vdash (x : A) \multimap_s B \utype L_k } 
      \rname{L-Prod}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype L_k \\ 
        \Gamma \vdash B \utype s_k }
      { \Gamma \vdash A \multimap_s B \utype L_k } 
      \rname{Lolli}
    \end{mathpar}
    \label{type}
  \end{figure}

>>>>

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{10pt}
\setlength{\grammarindent}{10em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\LNLD}{\text{LNL}_\text{D}}
\newcommand{\ok}{\textbf{ok}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{$\text{let }#1\text{ := }#2\text{ in }#3$}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\subst}{\text{subst}}
\newcommand{\utype}{:_U}
\newcommand{\ltype}{:_L}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\arw}[2]
{\hspace*{0.2em}{\scriptstyle #1}\hspace*{-0.2em}
\rightarrow
\hspace*{-0.2em}{\scriptstyle #2}\hspace*{0.2em}}

@@

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encodes the usage of resource. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (Money\ 5) \arw{L}{L} (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction states that it requires a payment of 5 units of money, the customer is relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{7.3em} expressions
      \indalt $(x : A) \arw{s}{t} B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \end{minipage}
    \label{syntax}
  \end{figure}
  A clear departure of our language from standard presentations of both linear type theory, and dependent type theory is the indexed function type: $(x : A) \arw{s}{t} B$. The reason for these indices is that we have built the ! exponential of linear logic directly into universe sorts. The indices $s$, and $t$ annotate the domain, and co-domain's respective universe sort, hence annotating linearity. The behavior of ! is difficult to account for even in simple linear type theory. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into type formation rules, we implicitly limit ! to only be used canonically. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without needing extra machinery for manipulating exponential.
  \begin{figure}[H]
    \caption{Correspondence of CLC types and MELL implications}
    \begin{align}
      (\_ : A) \arw{U}{U} B \quad &\equiv \quad !A \multimap !B \\
      (\_ : A) \arw{U}{L} B \quad &\equiv \quad !A \multimap B \\
      (\_ : A) \arw{L}{U} B \quad &\equiv \quad A \multimap !B \\
      (\_ : A) \arw{L}{L} B \quad &\equiv \quad A \multimap B
    \end{align}
    \label{correspondence}
  \end{figure}
  Figure \ref{correspondence} illustrates the correspondence between CLC function types and Multiplicative Exponential Linear Logic (MELL) implications. MELL lacks counterparts for the cases (1), (2) if the co-domain $B$ is dependent on arguments of the domain $A$. We will discuss the detail of these function types in Section \ref{tyformation}.

@@

  \begin{definition}
    The cumulativity relation ($\preceq$) is the smallest binary relation over terms such that
    \begin{enumerate}
      \item $\preceq$ is a partial order with respect to definitional equality.
        \begin{enumerate}
          \item If $A \equiv B$, then $A \preceq B$.
          \item If $A \preceq B$ and $B \preceq A$, then $A \equiv B$.
          \item If $A \preceq B$ and $B \preceq C$, then $A \preceq B$.
        \end{enumerate}
      \item $U_* \preceq U_0 \preceq U_1 \preceq U_2 \preceq \cdots$
      \item $L_0 \preceq L_1 \preceq L_2 \preceq \cdots$
      \item If $A_1 \equiv A_2$ and $B_1 \preceq B_2$, then
        $(x : A_1) \arw{s}{t} B_1 \preceq (x : A_2) \arw{s}{t} B_2$
    \end{enumerate}
  \end{definition}

  Figure \ref{universe} illustrates the structure of our universe hierarchy. Each linear universe $L_k$ has $U_{k+1}$ as its type, allowing functions to freely quantify over linear \textit{types}. However, $L_k$ cumulates to $L_{k+1}$. These two parallel threads of cumulativity prevent linear types from being transported to the non-linear universe, and subsequently losing track of its occupants' linearity.

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype U_i \\ 
        \Gamma, x \utype A \vdash B \utype U_* }
      { \Gamma \vdash (x : A) \arw{U}{U} B \utype U_* } 
      \rname{U-Prop}
      \\

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype U_k \\ 
        \Gamma, x \utype A \vdash B \utype s_k }
      { \Gamma \vdash (x : A) \arw{U}{s} B \utype t_k } 
      \rname{U-Prod}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A \utype L_k \\ 
        \Gamma \vdash B \utype s_k }
      { \Gamma \vdash (\_ : A) \arw{L}{s} B \utype t_k } 
      \rname{L-Prod}
    \end{mathpar}
    \label{type}
  \end{figure}
